% Vākya-Vallarī paper — compile: pdflatex -> bibtex -> pdflatex -> pdflatex
\documentclass[11pt]{article}

\usepackage[margin=1.05in]{geometry}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lmodern}
\usepackage{microtype}
\usepackage{natbib}
\usepackage{booktabs}
\usepackage{graphicx}
\usepackage{tikz}
\usepackage{pgfplots}
\pgfplotsset{compat=1.18}
\usetikzlibrary{shapes.geometric, arrows.meta, positioning, fit, calc, backgrounds, matrix}
\usepackage{listings}
\usepackage{amsmath, amssymb, amsthm}
\usepackage{xcolor}
\usepackage[colorlinks=true, linkcolor=blue!50!black, citecolor=blue!50!black, urlcolor=blue!50!black]{hyperref}

% IAST diacritics under pdflatex
\DeclareUnicodeCharacter{0101}{\=a}   % ā
\DeclareUnicodeCharacter{012B}{\={\i}}% ī
\DeclareUnicodeCharacter{016B}{\=u}   % ū
\DeclareUnicodeCharacter{1E5B}{\d{r}} % ṛ
\DeclareUnicodeCharacter{1E63}{\d{s}} % ṣ
\DeclareUnicodeCharacter{1E6D}{\d{t}} % ṭ
\DeclareUnicodeCharacter{1E0D}{\d{d}} % ḍ
\DeclareUnicodeCharacter{1E47}{\d{n}} % ṇ
\DeclareUnicodeCharacter{1E25}{\d{h}} % ḥ
\DeclareUnicodeCharacter{1E43}{\d{m}} % ṃ
\DeclareUnicodeCharacter{015B}{\'s}   % ś
\DeclareUnicodeCharacter{00F1}{\~n}   % ñ
\DeclareUnicodeCharacter{1E45}{\.n}   % ṅ

\definecolor{leanbg}{RGB}{246,241,231}
\definecolor{gatecol}{RGB}{180,60,50}
\definecolor{datacol}{RGB}{60,90,160}
\definecolor{proccol}{RGB}{60,130,80}
\lstdefinelanguage{lean}{
  keywords={theorem, def, structure, inductive, namespace, end, import, open, by, decide, deriving, where, example},
  sensitive=true, comment=[l]{--},
}
\lstset{basicstyle=\small\ttfamily, keywordstyle=\bfseries, commentstyle=\itshape\color{gray},
        frame=single, backgroundcolor=\color{leanbg!50}, breaklines=true, captionpos=b,
        literate={ā}{{\=a}}1 {ī}{{\={\i}}}1 {ṛ}{{\d{r}}}1 {ṣ}{{\d{s}}}1 {ṭ}{{\d{t}}}1
                 {ḍ}{{\d{d}}}1 {ṇ}{{\d{n}}}1 {ḥ}{{\d{h}}}1 {ṃ}{{\d{m}}}1 {ś}{{\'s}}1
                 {≠}{{$\neq$}}1 {¬}{{$\neg$}}1 {⟨}{{$\langle$}}1 {⟩}{{$\rangle$}}1 {→}{{$\to$}}1
                 {∧}{{$\wedge$}}1 {∨}{{$\vee$}}1 {│}{{|}}1}

\newtheorem{definition}{Definition}

\title{Machine-Checked Translation Adequacy for the Brahmak\={a}\d{n}\d{d}a of
  Bhart\d{r}hari's \emph{V\={a}kyapad\={\i}ya}:\\ Commentary-Driven Verification
  of a Sanskrit Philosophical Text in Lean~4}

\author{Sharath Sathish\\
  \small Independent Researcher\\
  \small \texttt{sharath.sathish@gmail.com}\\
  \small \url{https://github.com/SharathSPhD/vakya-vallari}}

\date{July 2026}

\begin{document}
\maketitle

\begin{abstract}
Translations of dense philosophical Sanskrit are under-determined. A single
k\={a}rik\={a} admits several grammatically valid renderings whose
philosophical commitments differ radically, and the commentarial tradition,
not surface grammar, fixes the intended reading. This paper introduces
commentary-driven translation validation (C-TV), an adaptation of translation
validation from compiler verification to natural language, and applies it to
the complete Brahmak\={a}\d{n}\d{d}a (Book~I, 144 verse units) of
Bhart\d{r}hari's \emph{V\={a}kyapad\={\i}ya}. A machine-produced English
translation and analytical commentary, prepared with K.~A. Subramania Iyer's
edition and translation of Book~I with the V\d{r}tti as the primary reference
text, is compiled verse by verse into a semantic contract: sorted entities,
identity, relation, and predication claims, and explicit denials, with every
axiom citing the commentary verbatim. A generator compiles each contract into
Lean~4, where a decidable adequacy predicate proves that the accepted
translation satisfies the contract (144 adequacy theorems) and that 361
plausible mistranslations provably fail (156 contradicted by commentary
denials, 205 unlicensed by the axioms), for 566 machine-checked theorems with
zero \texttt{sorry}. The contracts were authored by an agent swarm gated by a
validator and by the Lean kernel; the gates caught errors at every scale,
including fabricated citations in quota-degraded agent output and
cross-contract ontological drift, which adjudication resolved into a registry
of 41 polysemous terms with 91 textually justified senses. A structured
adversarial review (\S\ref{sec:adversarial}) stress-tested the method: most
attacks failed, a cross-verse swap test found zero of 20{,}592 pairs
interchangeable, and the one major finding---that a verbatim-citation gate
cannot distinguish endorsed doctrine from reported p\={u}rvapak\d{s}a---led to
an explicit doxographic-stance mechanism. An embedding audit of
translation--commentary pairs provides a ranked review queue and is
reported with its limits. The verification boundary is stated precisely: Lean
checks the internal coherence of the formalization against the
commentary-derived contract, and the contract--commentary link remains
human-auditable through the verbatim citations. All artifacts, including the
corpus, contracts, kernel, proofs, and edition, are public and reproducible.
\end{abstract}

\section{Introduction}
\label{sec:intro}

The \emph{V\={a}kyapad\={\i}ya} (Treatise on Sentence and Word, c.~5th
century CE) is the foundational text of Indian philosophy of language. Its
opening verse already exhibits the problem this paper addresses:

\begin{quote}
\emph{an\={a}dinidhana\d{m} brahma \'sabdatattva\d{m} yadak\d{s}aram\,/\\
vivartate'rthabh\={a}vena prakriy\={a} jagato yata\d{h}\,//} (VP 1.1)
\end{quote}

A translator must decide whether \emph{\'sabdatattva} (word-essence) is a
property of language or the Absolute itself; whether \emph{vivartate} means
that the world is a transformation of Brahman (pari\d{n}\={a}ma) or its
appearance without transformation (vivarta); and whether \emph{ak\d{s}ara}
means imperishable, phoneme, or, as the grammarian tradition insists,
deliberately both. Each choice yields a grammatical English sentence. The
choices differ philosophically as much as materialism differs from idealism.
What disambiguates them is not syntax but the commentarial apparatus of
v\d{r}tti and \d{t}\={\i}k\={a} that fixes, verse by verse, which reading the
text licenses \citep{iyer1965vakyapadiya, houben2016vakyapadiya}.

Translation of such texts has therefore been a discipline whose constraints,
though real, live in the philologist's head. This paper makes a subset of
those constraints mechanical. The claim is not that a theorem prover can read
Sanskrit. The claim is that once interpretive commitments are stated as data,
a proof assistant can hold a translation to them exhaustively, incrementally,
and publicly.

\paragraph{Scope.} The verified object is the Brahmak\={a}\d{n}\d{d}a, the
first of the text's three books, in its entirety: 144 verse units, of which
56 carry contested readings. Book~I contains the text's core metaphysics of
language (\'sabd\={a}dvaita and the spho\d{t}a doctrine) and is the book for
which \citet{iyer1965vakyapadiya} provides the edition with the V\d{r}tti
that served as the primary reference for the translation and commentary
formalized here. Books~II and~III exist in the released corpus as unverified
context and are outside the scope of the present claims.

\paragraph{Contributions.}
\begin{enumerate}
  \item \textbf{Commentary-driven translation validation (C-TV).} The paper
    adapts translation validation from compiler verification
    \citep{pnueli1998translation, leroy2009compcert, tristan2008validators}
    to natural language. The commentary is compiled into per-verse semantic
    contracts (\S\ref{sec:method}), and a translation's reading is verified
    against the contract rather than against the bare source string. To the
    author's knowledge this is the first application of an interactive
    theorem prover to translation adequacy in linguistics, and the first
    mechanized treatment of a Sanskrit philosophical
    text.\footnote{Prior mechanizations of Indian logic formalize the logical
    systems (\S\ref{sec:related}); none, to the author's knowledge, verifies
    translations of texts.}
  \item \textbf{A verified Brahmak\={a}\d{n}\d{d}a.} All 144 verse units of
    Book~I carry compiled adequacy theorems, and 361 plausible
    mistranslations are compiled counterexamples, refuted by theorem rather
    than by footnote (\S\ref{sec:kernel}, \S\ref{sec:results}).
  \item \textbf{An anti-fabrication architecture for agentic formalization.}
    The contracts were authored by an LLM agent swarm, and every layer is
    gated: axioms must cite the commentary verbatim; rejected readings must
    provably fail; entity sorts must be globally consistent or adjudicated as
    polysemy in a justified registry (\S\ref{sec:loop},
    \S\ref{sec:ontology}). The paper reports what the gates caught, including
    fabricated citations in quota-degraded agent output, as evidence that the
    architecture rather than agent reliability carries the epistemic load.
  \item \textbf{A ranked semantic review queue.} An embedding audit of
    translation--commentary pairs yields a similarity-ranked queue for
    editorial attention, reported together with a negative finding about its
    discriminative power at book scale (\S\ref{sec:audit}).
\end{enumerate}

\begin{figure}[t]
\centering
\begin{tikzpicture}[
  node distance=3.6mm and 6mm,
  box/.style={draw, rounded corners=2pt, align=center, font=\small,
              inner sep=4pt, minimum height=8mm},
  gate/.style={box, fill=red!8, draw=gatecol},
  data/.style={box, fill=blue!6, draw=datacol},
  proc/.style={box, fill=green!7, draw=proccol},
  arr/.style={-{Stealth[length=2.2mm]}, semithick}]
\node[data] (iyer) {Iyer 1965\\ (primary text)};
\node[data, right=of iyer] (mula) {m\={u}la\\ (Devanagari)};
\node[proc, right=of mula] (trans) {translation +\\ commentary};
\node[proc, below=of $(mula.south)!0.5!(trans.south)$] (corpus)
  {canonical corpus\\ Book~I: 144 units};
\node[gate, right=of corpus] (g1) {corpus gates\\ (pytest)};
\node[proc, below=of corpus] (swarm) {agent swarm\\ drafts contracts};
\node[gate, right=of swarm] (g2) {validator: verbatim cites,\\ entailment, refutability};
\node[proc, below=of swarm] (leangen) {generator\\ $\to$ Lean modules};
\node[gate, right=of leangen] (g3) {\texttt{lake build},\\ zero \texttt{sorry}};
\node[proc, below=of leangen] (adj) {consistency gate +\\ homonym registry};
\node[gate, right=of adj] (g4) {adjudication\\ (textual merits)};
\node[data, below=of adj] (out) {verified edition\\ 144 adequacy + 361 refutation thms};
\draw[arr] (iyer) -- (trans);
\draw[arr] (mula) -- (corpus);
\draw[arr] (trans) -- (corpus);
\draw[arr] (corpus) -- (g1);
\draw[arr] (corpus) -- (swarm);
\draw[arr] (swarm) -- (g2);
\draw[arr] (g2.south) to[bend left=18] (swarm.east);
\draw[arr] (swarm) -- (leangen);
\draw[arr] (leangen) -- (g3);
\draw[arr] (g3.south) to[bend left=18] (leangen.east);
\draw[arr] (leangen) -- (adj);
\draw[arr] (adj) -- (g4);
\draw[arr] (g4.south) to[bend left=18] (adj.east);
\draw[arr] (adj) -- (out);
\end{tikzpicture}
\caption{The C-TV pipeline. Every layer has a mechanical gate (red). Failures
feed back as repair work, and an append-only research ledger records each
draft, rejection, repair, and acceptance.}
\label{fig:pipeline}
\end{figure}

\section{Background and Related Work}
\label{sec:related}

\subsection{Bhart\d{r}hari and the Brahmak\={a}\d{n}\d{d}a}
The \emph{V\={a}kyapad\={\i}ya} comprises three books
(Brahmak\={a}\d{n}\d{d}a, V\={a}kyak\={a}\d{n}\d{d}a, Padak\={a}\d{n}\d{d}a).
Book~I develops \'sabd\={a}dvaita, the thesis that the Absolute is of the
nature of the word, and the spho\d{t}a doctrine: the meaning-bearing word is
an undivided whole, merely manifested by the temporal sequence of sounds
\citep{coward1980sphota, biardeau1964theorie, staal1969sanskrit}. Both
doctrines generate exactly the translation hazards the present system checks:
sort errors (demoting the Absolute to a property), relation errors (rendering
manifestation as transformation), and homonym errors (collapsing the
technical senses of ak\d{s}ara, spho\d{t}a, and dhvani). The standard modern
edition of Book~I with the V\d{r}tti and an English translation is
\citet{iyer1965vakyapadiya}; \citet{houben1995sambandha} treats
Bhart\d{r}hari's theory of relation, which motivates the kernel's nested
relations (\S\ref{sec:kernel}).

\subsection{Computational Sanskrit}
P\={a}\d{n}inian grammar has long been recognized as a formal generative
system \citep{kiparsky2009panini, cardona1976panini, scharf2009panini}, and a
mature computational ecosystem exists for segmentation, morphology, and
generation \citep{goyal2016samsaadhanii, hellwig2020sanskrit, vidyut}, with
TEI-encoded editions in SARIT \citep{sarit}. That work is orthogonal and
complementary: the present system verifies semantics against commentary, not
surface analysis.

\subsection{Formalizing Indian logic}
Navya-Ny\={a}ya has been studied logically since \citet{ingalls1951materials}
and \citet{matilal1968navya}. Recently \citet{panday2026cubical} encoded its
core constructs (relational nesting, delimitors, typed absence) in cubical
type theory, \citet{sathish2026pramana} trained LLMs on its six-phase
epistemic method, and \citet{diehl2024fourfold} mechanized the Buddhist
catu\d{s}ko\d{t}i in Lean. These works formalize the logics; none verifies
translations of a text. The kernel here borrows Navya-Ny\={a}ya's structural
insights in plain Lean~4 \citep{demoura2021lean4}, without cubical machinery.

\subsection{Translation validation and autoformalization}
Translation validation checks each compiler run a posteriori instead of
verifying the compiler \citep{pnueli1998translation, tristan2008validators}.
The present system transplants the pattern: the translator is the untrusted
optimizer, the contract is the semantic reference, and Lean is the validator.
On the authoring side, LLM autoformalization \citep{wu2022autoformalization,
pathak2024gflean} supplies the untrusted draft whose output the kernel gates.
The loop is an instance of the generate-and-check paradigm with a strict
checker.

\subsection{Gate-audited agentic research programmes}
The formalization loop follows a discipline developed in two companion
programmes. Active Circuit Discovery frames mechanistic interpretability
under a budget as an experimental-design problem and delegates experiment
selection to an active-inference agent \citep{sathish2026acd}. The Prabodha
programme runs a pre-registered, gate-audited research loop in which every
claim must pass both a code gate and a domain gate, with an append-only
evidence ledger and adversarial review \citep{sathish2026prabodha}. The
present system inherits the pattern: agents propose, mechanical gates
dispose, and an append-only ledger records every draft, rejection, repair,
and acceptance. The difference is the checker. Where those programmes gate on
statistical criteria, this one gates on a type-checker, so acceptance is a
theorem rather than a threshold.

\section{Method: Commentary-Driven Translation Validation}
\label{sec:method}

\subsection{Problem statement}
Let $v$ be a verse with source text $s_v$, commentary $c_v$, and candidate
translation $t_v$. Direct verification of $t_v$ against $s_v$ is not well
posed: the source under-determines the reading, and the commentary is what
fixes it. C-TV therefore verifies $t_v$ against a formal object derived from
$c_v$.

\begin{definition}[Semantic contract]
A contract for verse $v$ is a JSON document with (i) \emph{entities}: named
terms with ontological sorts drawn from $\{$absolute, power, manifestation,
linguisticItem, property, cognition$\}$; (ii) \emph{axioms} and
\emph{denials}: identity, relation, and predication claims over those
entities, each carrying a citation that must occur verbatim in $c_v$;
(iii) an \emph{accepted reading}: the claims that $t_v$ asserts, each citing
$t_v$ verbatim; and (iv) \emph{rejected readings}: plausible mistranslations
rendered as claim sets, each annotated with the failure mode it must
exhibit.
\end{definition}

\begin{definition}[Adequacy and failure modes]
A reading $r$ is \emph{adequate} for contract $C$ when every claim of $r$ is
entailed by the axioms of $C$ and no claim of $r$ is denied by $C$. A
rejected reading must fail in one of two registered modes:
\texttt{contradicted} (it asserts a denial) or \texttt{unlicensed} (it
asserts a claim no axiom entails). Both modes are decidable
(\S\ref{sec:kernel}).
\end{definition}

\subsection{The anti-fabrication gate}
The verbatim-citation rule is the load-bearing constraint. An axiom that
cannot point to the commentary text that licenses it is rejected
mechanically, before any proof is attempted. This converts the usual
free-form risk of LLM-authored formal content (invented support) into a
string-matching check. The same rule applies to the accepted reading, whose
claims must cite the translation verbatim, so the contract cannot silently
attribute to the translation a claim it does not make.

\subsection{Worked example: VP 1.1}
\label{sec:worked}
Figure~\ref{fig:readings} shows the compiled fate of three candidate
renderings of the opening verse. The contract for 1.1 declares three
entities (brahman and \'sabdatattva of sort absolute, jagat of sort
manifestation), four axioms (the identity of \'sabdatattva and brahman, the
vivarta relation to the world, and the predications an\={a}dinidhana and
ak\d{s}ara), and one denial (the pari\d{n}\={a}ma relation). The accepted
rendering asserts four claims, each citing the translation text. Two
rejected renderings are registered. The first demotes \'sabdatattva to a
property of language; its identity claim then crosses sorts and no axiom
entails it, so it fails as \texttt{unlicensed}, with an auxiliary theorem
recording the sort error. The second renders vivartate as transformation; it
asserts exactly the denied pari\d{n}\={a}ma relation and fails as
\texttt{contradicted}.

\begin{figure}[t]
\centering
\begin{tikzpicture}[
  node distance=4.5mm and 5mm,
  cand/.style={draw, rounded corners=2pt, align=left, font=\small, inner sep=4pt,
               fill=blue!5, draw=datacol, text width=56mm},
  verdict/.style={draw, rounded corners=2pt, align=center, font=\small, inner sep=4pt},
  ok/.style={verdict, fill=green!10, draw=proccol},
  bad/.style={verdict, fill=red!8, draw=gatecol},
  arr/.style={-{Stealth[length=2.2mm]}, semithick}]
\node[cand] (a) {``That beginningless and endless One, Brahman, whose essence
  is the Word, imperishable, appears in the mode of objects\ldots''};
\node[cand, below=of a] (b) {``The imperishable linguistic structure
  underlying words has no beginning or end.''};
\node[cand, below=of b] (c) {``Brahman transforms itself into the world of
  objects.''};
\node[ok, right=14mm of a] (va) {\texttt{Adequate}\\ theorem by \texttt{decide}};
\node[bad, right=14mm of b] (vb) {\texttt{unlicensed}\\ sort error:\\
  property $\neq$ absolute};
\node[bad, right=14mm of c] (vc) {\texttt{contradicted}\\ asserts denied\\
  pari\d{n}\={a}ma relation};
\draw[arr] (a) -- (va);
\draw[arr] (b) -- (vb);
\draw[arr] (c) -- (vc);
\node[font=\small\itshape, above=2mm of a] {candidate renderings of VP 1.1};
\node[font=\small\itshape, above=2mm of va] {compiled verdict};
\end{tikzpicture}
\caption{Three renderings of VP 1.1 and their compiled verdicts. The
accepted reading is proved adequate; each rejected reading is refuted by a
theorem in its registered failure mode.}
\label{fig:readings}
\end{figure}

\subsection{Verification boundary}
\label{sec:boundary}
Lean verifies the internal coherence of the formalization: the accepted
reading satisfies the contract and the rivals provably do not. Whether the
contract is faithful to Bhart\d{r}hari is not a theorem and cannot be one.
That link is kept auditable by construction: every axiom carries the exact
commentary sentence it derives from, so checking a contract against the
commentary is a lookup rather than a reconstruction. The division of labor
is deliberate. The prover keeps commitments; the philologist judges them.

\section{The Corpus}
\label{sec:corpus}

\subsection{Sources}
The edition is built from two sources. The Devanagari m\={u}la
k\={a}rik\={a}s were scraped from a public edition (wisdomlib.org). The
English translation and analytical commentary were produced afresh for this
project by a machine-assisted process working from the Sanskrit, with the
edition and annotated translation of Book~I with the V\d{r}tti by
\citet{iyer1965vakyapadiya} as the primary reference text. Interpretive
decisions in the commentary follow the V\d{r}tti as presented there;
disputed readings are recorded in a per-verse \emph{contested} note rather
than silently resolved.

\subsection{Alignment and gates}
Alignment between m\={u}la and commentary is checked, never assumed. The
commentary units are the canonical keys; every unit must match a m\={u}la
entry by exact label or be quarantined with a reason. The build is
test-driven, and the tests forced three philologically real discoveries: the
\emph{contested} field is a prose note explaining each disputed reading, not
a boolean; verse markers legitimately embed Latin half-verse letters; and the
source edition assigns one label to two different pages, a defect the
commentary itself documents, which the loader merges and flags rather than
silently overwriting. The resulting JSONL corpus carries per-record
provenance and a manifest with a SHA-256 checksum.
Table~\ref{tab:corpus} summarizes the corpus and the verified subset.

\begin{table}[t]
\centering
\caption{The corpus and the verified scope. Books~II and~III are released as
unverified context; all verification claims in this paper concern Book~I.}
\label{tab:corpus}
\begin{tabular}{lrrrr}
\toprule
 & \textbf{Book I} & \textbf{Book II} & \textbf{Book III} & \textbf{Total} \\
\midrule
Verse units (canonical)          & 144 & 426 & 1,226 & 1,796 \\
Contested readings (prose notes) & 56  & 167 & 63    & 286 \\
Semantic contracts               & 144 & --- & ---   & 144 \\
Adequacy theorems (Lean)         & 144 & --- & ---   & 144 \\
Refutation theorems (Lean)       & 361 & --- & ---   & 361 \\
\bottomrule
\end{tabular}
\end{table}

\section{Contract Anatomy and Statistics}
\label{sec:contracts}

Across the 144 contracts there are 776 entity declarations, 624 axioms (of
which 32, across nine doxographic verses, carry a \emph{reported} stance;
\S\ref{sec:kernel}), 181 denials, 450 accepted-reading claims, and 361
rejected readings. Every contract carries at least two rejected readings (76
carry two, 63 carry three, five carry four), a floor imposed after the
adversarial evaluation of \S\ref{sec:adversarial}. Figure~\ref{fig:stats}
shows the distribution of entity sorts and axiom kinds. Relations dominate
the axioms (320 of 624), which reflects the text: Book~I argues chiefly
about how things stand to one another (manifestation, expression,
dependence), not about what things are called.

The validator checks, per contract: schema validity; sort membership; ASCII
identifier hygiene (a Lean requirement discovered during the loop,
\S\ref{sec:loop}); verbatim presence of every citation; stance well-formedness;
disjointness of axioms and denials (no claim may be both licensed and denied);
at least two rejected readings; entailment of the accepted reading; and
provable failure of every rejected reading in its registered mode. The
entailment and refutation checks run the same decision procedure the Lean
kernel uses, so Python-side acceptance predicts Lean-side theoremhood.

\begin{figure}[t]
\centering
\begin{tikzpicture}
\begin{axis}[width=0.46\textwidth, height=5.2cm, ybar,
  bar width=7pt, xtick=data, x tick label style={rotate=35, anchor=east, font=\footnotesize},
  symbolic x coords={linguisticItem, manifestation, property, power, cognition, absolute},
  ylabel={entities}, label style={font=\small}, tick label style={font=\small},
  ymin=0, enlarge x limits=0.12, title={\small entity sorts (776)}]
\addplot+[fill=blue!25, draw=datacol] coordinates {
 (linguisticItem,192) (manifestation,149) (property,134) (power,126)
 (cognition,90) (absolute,85)};
\end{axis}
\end{tikzpicture}\hfill
\begin{tikzpicture}
\begin{axis}[width=0.46\textwidth, height=5.2cm, ybar,
  bar width=10pt, xtick=data, x tick label style={font=\footnotesize},
  symbolic x coords={relation, predication, identity, denial},
  ylabel={claims}, label style={font=\small}, tick label style={font=\small},
  ymin=0, enlarge x limits=0.2, title={\small axiom kinds (624) and denials}]
\addplot+[fill=green!25, draw=proccol] coordinates {
 (relation,320) (predication,273) (identity,31) (denial,181)};
\end{axis}
\end{tikzpicture}
\caption{Contract statistics over the 144 Brahmak\={a}\d{n}\d{d}a contracts.
Left: entity declarations by ontological sort. Right: axioms by kind, with
denials shown alongside.}
\label{fig:stats}
\end{figure}

\section{The Lean Kernel}
\label{sec:kernel}

\subsection{Design}
The kernel (\texttt{VakyaVallari}, Lean~4.32, no external dependencies) is
deliberately small: the ontology and adequacy core is under 150 lines.
Entities are sort-tagged terms; claims are inductive data; a contract is
axioms plus denials; and adequacy is decidable, so every verse-level theorem
closes by \texttt{decide}. The proof term is the computation.

\begin{lstlisting}[language=lean, caption={Core of the adequacy kernel (abridged).}]
inductive Sorta | absolute | power | manifestation
               | linguisticItem | property | cognition
structure Entity where  name : String;  sort : Sorta
-- paramparā: a relation can itself be a relatum, to any depth
inductive Node | ent : Entity → Node
              | rel : String → Node → Node → Node
structure Abhava where  pratiyogin : Node;  adhikarana : Node
structure Contract where  axioms : List Claim;  denials : List Claim
                         reported : List Claim := []   -- purvapaksa
def Contract.doxographic (c : Contract) : Bool := !c.reported.isEmpty
def Adequate (c : Contract) (r : Reading) : Prop :=
  r.claims.all (entails (c.axioms ++ c.reported))
    ∧ ¬ r.claims.any (denied c.denials)
\end{lstlisting}

\subsection{Navya-Ny\={a}ya structure}
Three design choices are drawn from Navya-Ny\={a}ya analysis
\citep{ingalls1951materials, matilal1968navya}. \emph{Sorted identity}:
identity claims are stateable across sorts but decidably false when sorts
differ, so a mistranslation that demotes the Absolute to a property fails by
sort error, the C-TV failure mode realized as a computation. \emph{Nested
relations}: \texttt{Node} makes relations first-class relata
(parampar\={a}-sambandha, cf.~\citealp{houben1995sambandha}), so a claim
such as ``time is the instrumental cause of the manifestation-relation'' is
representable without ad hoc reification. \emph{Typed absence}:
\texttt{Abhava} carries its counterpositive and locus as data, and the
kernel proves the Navya-Ny\={a}ya involution (the absence of an absence
computes to the presence) as a definitional fact. We are precise about this
last item's status: \texttt{Abhava} and the involution are exercised by the
kernel's own tests (\texttt{Tests.lean}), not by any verse module in the
present corpus. They are provided machinery that the negation-heavy verses of
Books~II--III are expected to use, not load-bearing in the Book-I results;
the Book-I refutations rest on sorted identity and relation matching.

\subsection{Doxographic stance}
\label{sec:stance}
A verbatim-citation gate proves a cite \emph{occurs} in the commentary; it
cannot tell an endorsed doctrine from a reported rival view
(p\={u}rvapak\d{s}a), since both are present in the prose. Book~I contains a
doxographic stretch---the phonetician theories of sound (1.102--1.106), the
substrate doctrines (1.107--1.113), and the position-reporting verses
1.68--69, 1.70, 1.81---where a naive contract would encode ``some say $X$'' as
an endorsed axiom. The kernel therefore separates \texttt{reported} claims
from \texttt{axioms}. Reported claims still \emph{license} a reading, because a
faithful translation of a doxographic verse must be free to assert the report;
but the split is machine-visible (\texttt{Contract.doxographic}), the validator
requires reported axioms to be marked, and the edition surfaces the distinction
per verse. Nine contracts carry 32 reported axioms. This does not make the
system able to \emph{judge} endorsement---that remains the human's job, audited
through the citation---but it removes the false suggestion that a doxographic
report is Bhart\d{r}hari's own commitment.

\begin{figure}[t]
\centering
\begin{tikzpicture}[
  node distance=4mm and 8mm,
  ty/.style={draw, rounded corners=2pt, align=center, font=\small\ttfamily,
             inner sep=4pt, fill=blue!5, draw=datacol},
  lbl/.style={font=\footnotesize\itshape, text=black!70},
  arr/.style={-{Stealth[length=2.2mm]}, semithick}]
\node[ty] (sorta) {Sorta\\ (6 sorts)};
\node[ty, right=of sorta] (ent) {Entity\\ name $\times$ sort};
\node[ty, right=of ent] (node) {Node\\ ent $\mid$ rel};
\node[ty, right=of node] (abh) {Abhava\\ pratiyogin $\times$\\ adhikarana};
\node[ty, below=of ent] (claim) {Claim\\ identity $\mid$ relation $\mid$\\ predication};
\node[ty, right=of claim] (contract) {Contract\\ axioms $\times$ denials};
\node[ty, right=of contract] (reading) {Reading\\ claims};
\node[ty, below=of contract] (adeq) {Adequate c r : Prop\\ (decidable)};
\draw[arr] (sorta) -- (ent);
\draw[arr] (ent) -- (node);
\draw[arr] (node) -- (abh);
\draw[arr] (ent) -- (claim);
\draw[arr] (node) to[bend right=12] (claim);
\draw[arr] (claim) -- (contract);
\draw[arr] (claim) -- (reading);
\draw[arr] (contract) -- (adeq);
\draw[arr] (reading) -- (adeq);
\node[lbl, below=1mm of node] {nesting depth unbounded};
\node[lbl, below=1mm of adeq] {closed by \texttt{decide}};
\end{tikzpicture}
\caption{Type structure of the kernel. Relations are first-class relata via
\texttt{Node}; absence is typed data; adequacy is a decidable proposition
over a reading and a contract.}
\label{fig:kernel}
\end{figure}

\subsection{Generated modules}
Each contract compiles to a Lean module in which the accepted reading's
adequacy is a theorem and every rejected reading's failure is a theorem.
Sort-error refutations additionally carry an auxiliary lemma naming the
mismatch; there are 56 such lemmas across the corpus. Verse 1.1's module
ends as follows:

\begin{lstlisting}[language=lean, caption={Generated theorems for VP 1.1 (excerpt).}]
theorem accepted_adequate : contract.Adequate accepted := by decide
-- "The imperishable linguistic structure underlying words..."
theorem naive_linguistic_structure_inadequate :
    ¬ contract.Adequate naive_linguistic_structure := by decide
theorem naive_linguistic_structure_sort_error :
    sabdatattva_naive ≠ sabdatattva := by decide
-- "Brahman transforms itself into the world of objects."
theorem parinama_transformation_inadequate :
    ¬ contract.Adequate parinama_transformation := by decide
\end{lstlisting}

The build gate is binary and global: \texttt{lake build} must succeed with
zero \texttt{sorry} across all 144 modules, 566 theorems at present. A
counterexample that stops compiling is a regression, caught in CI.

\section{The Agentic Formalization Loop}
\label{sec:loop}

\subsection{Architecture}
Contracts were authored by an LLM agent swarm: one agent per verse, each
given the verse record, the schema, an exemplar, and the validator as its
exit gate. Rounds of 9 to 30 agents ran with an adjudication pass between
rounds and an append-only ledger recording every event (96 events: 29
drafts, 40 acceptances, 9 gate runs, 6 validations, 3 rejections, 3 repairs,
5 milestones, 1 publication). The division of labor is the point: agents
propose; gates dispose. Figure~\ref{fig:loop} shows one round.

\begin{figure}[t]
\centering
\begin{tikzpicture}[
  box/.style={draw, rounded corners=2pt, align=center, font=\small,
              inner sep=4pt, minimum height=9mm},
  gate/.style={box, fill=red!8, draw=gatecol},
  proc/.style={box, fill=green!7, draw=proccol},
  data/.style={box, fill=blue!6, draw=datacol},
  arr/.style={-{Stealth[length=2.2mm]}, semithick},
  lab/.style={font=\footnotesize, fill=white, inner sep=1pt}]
\node[data] (queue) at (0,0) {verse queue\\ (Book I)};
\node[proc] (agent) at (3.2,0) {per-verse agent\\ drafts contract};
\node[gate] (val) at (7.0,0) {validator: schema, cites,\\ entailment, refutability};
\node[gate] (lean) at (11.0,0) {generator +\\ \texttt{lake build}};
\node[proc] (repair) at (3.2,-2.2) {repair agent\\ (bounded retries)};
\node[proc] (adj) at (11.0,-2.2) {adjudication:\\ consistency gate,\\ homonym registry};
\node[data] (ledger) at (7.0,-3.8) {append-only ledger (96 events)};
\draw[arr] (queue) -- (agent);
\draw[arr] (agent) -- node[lab, above]{draft} (val);
\draw[arr] (val) -- node[lab, above]{pass} (lean);
\draw[arr] (lean) -- node[lab, right]{green, zero sorry} (adj);
\draw[arr, rounded corners=3pt] (val.south) -- (7.0,-1.1) --
  node[lab, above]{fail} (3.2,-1.1) -- (repair.north);
\draw[arr, rounded corners=3pt] (lean.south) -- (11.0,-1.0) -- (9.2,-1.0)
  -- (4.2,-1.6) -- (repair.north east);
\draw[arr] (repair) -- node[lab, left]{redraft} (agent);
\draw[arr, rounded corners=3pt] (adj.south) |- (ledger.east);
\draw[arr, rounded corners=3pt] (ledger.west) -- (0,-3.8) --
  node[lab, left]{next verse} (queue.south);
\end{tikzpicture}
\caption{One round of the formalization loop. Every acceptance passes the
validator, the Lean kernel, and the cross-contract consistency gate; every
event is appended to the ledger.}
\label{fig:loop}
\end{figure}

\subsection{What the gates caught}
Four episodes illustrate the gate stack functioning under real failure.
Table~\ref{tab:catches} summarizes.

\begin{table}[t]
\centering
\caption{Errors caught mechanically during formalization. A lazy
counterexample is a rejected reading whose claims were secretly entailed; a
fabricated citation is an axiom cite absent from the commentary.}
\label{tab:catches}
\begin{tabular}{lrl}
\toprule
\textbf{Failure class} & \textbf{Count} & \textbf{Caught by} \\
\midrule
Lazy counterexamples             & 3 & Lean kernel, then hardened validator \\
Fabricated / paraphrased citations & 5 & verbatim-cite gate \\
Non-ASCII Lean identifiers       & 2 contracts & Lean parser, then validator \\
Cross-contract sort drift        & 95 conflicts & consistency gate \\
Diacritic name splits (spho\d{t}a/sphota) & 2 episodes & consistency gate \\
\bottomrule
\end{tabular}
\end{table}

\emph{(i) Lazy counterexamples.} Early agents produced rejected readings
that merely restated axioms. The corresponding inadequacy theorems would
have been false, and the Lean build failed. The validator was hardened to
require provable failure in the registered mode, and three contracts were
repaired.

\emph{(ii) Fabricated citations under degradation.} When an API session
limit killed 20 of 30 agents mid-round, 9 left unreviewed draft contracts.
Five of the nine failed the verbatim-citation gate with paraphrased or
invented cites. This is the anti-fabrication architecture functioning as
designed: degraded generation was caught by the gate, quarantined, and
re-authored.

\emph{(iii) Identifier hygiene.} Lean's parser rejected IAST diacritics in
entity identifiers that Python's \texttt{isidentifier()} accepts. The
validator now enforces ASCII identifiers. The incident argues for keeping
the end-to-end kernel gate even when a faster mirror exists.

\emph{(iv) Sort drift.} Independent agents assigned different sorts to the
same term. Across six adjudication rounds the consistency gate surfaced 95
conflicts (some terms recurring across rounds as new verses entered);
adjudication resolved each as either a correction or genuine polysemy,
converging on the registry of \S\ref{sec:ontology}. The per-round counts are
recorded in the research ledger.

\section{Ontological Consistency and the Homonym Registry}
\label{sec:ontology}

The corpus-level gate requires that a named entity carry one sort everywhere,
unless the exact partition of (verse, sort) uses is registered as polysemy
with a gloss and a one-sentence textual justification per sense. The
registry holds 41 terms with 91 senses. This is where the method earns its
keep philologically: the gate cannot tell drift from doctrine, so every
split is a documented editorial ruling.

Three registered splits are load-bearing. \emph{ak\d{s}ara}: the
imperishable Absolute in 1.9 versus phoneme in 1.18--22. This is the pun
that the commentarial tradition treats as the argument of the opening verse
in miniature; a same-sort rule would have flattened it. \emph{\'sabda}:
linguisticItem canonically, but the Absolute in the fire-in-wood verse
(1.46), where the inner word is \'sabdatattva itself. \emph{spho\d{t}a}:
linguisticItem in Bhart\d{r}hari's own doctrine, but verses 1.102--1.106
report a rival phonetician terminology in which spho\d{t}a names the
first-produced acoustic sound. Registering that doxographic span as a
distinct sense (manifestation) keeps the report from contaminating the
doctrine. A further adjudication round registered polysemy for veda (the
recited corpus versus the eternal seed-order of 1.132), \'sruti,
vyavasth\={a}, r\={u}pa, s\={a}dhu, vipar\={\i}ta, and k\={a}rya.

\section{Results}
\label{sec:results}

\subsection{Verification results}
Table~\ref{tab:results} summarizes the verified state. CI reproduces it from
a clean checkout (corpus gates, \texttt{lake build}, zero-\texttt{sorry}
sweep).

\begin{table}[t]
\centering
\caption{Verification results for the complete Brahmak\={a}\d{n}\d{d}a.}
\label{tab:results}
\begin{tabular}{lr}
\toprule
Verse units formalized and verified & 144 / 144 \\
Entity declarations                & 776 \\
Contract axioms (endorsed / reported) & 592 / 32 \\
Denials                            & 181 \\
Doxographic contracts (carry reported axioms) & 9 \\
Accepted-reading claims (all proved adequate) & 450 \\
Rejected readings, \texttt{contradicted} (denial asserted) & 156 \\
Rejected readings, \texttt{unlicensed} (no axiom entails)  & 205 \\
Auxiliary sort-error lemmas        & 56 \\
Lean theorems (all by \texttt{decide}, zero \texttt{sorry}) & 566 \\
Homonym registry (terms / senses)  & 41 / 91 \\
Cross-licensed verse pairs (of 20{,}592) & 0 \\
Python gates (pytest)              & 190 \\
Ledger events (append-only)        & 99 \\
\bottomrule
\end{tabular}
\end{table}

\subsection{A map of how Book I can be mistranslated}
Beyond the counts, the qualitative yield is a machine-checked map of the
ways Book~I goes wrong in translation. Each of the 361 refutations is a
specific, named failure: pari\d{n}\={a}ma for vivarta (1.1); grammar demoted
from discipline to convention (1.11--1.14); inference granted an autonomy
the text denies it (1.30--1.42); spho\d{t}a temporalized (1.75--1.77); the
corrupt form elevated to equal standing with the correct (1.147--1.155).
Together the refutations form a negative image of the doctrine, in theorem
form. Because each is a compiled theorem, the map regresses: an edit to any
contract that would quietly license a refuted reading breaks the build.

\subsection{The edition}
The edition is deployed as a static site with per-verse pages carrying
Devanagari, IAST, translation, commentary, contested notes, and per-verse
proof status, with the repository, contracts, kernel, and ledger public.

\section{Embedding Audit as a Review Queue}
\label{sec:audit}

Formal verification establishes coherence between translation and contract.
A separate instrument screens for drift between translation and commentary
prose. Every verse's translation and commentary are embedded
(all-mpnet-base-v2 \citep{reimers2019sbert}, mean-pooled, on an NVIDIA
GB10) and verses are ranked by cosine similarity.

Within Book~I the distribution has mean 0.557
(Figure~\ref{fig:audit}). The audit is reported with a negative finding: at
book scale the contested flag does not separate. Contested verses average
0.567 against 0.550 for uncontested verses, and the lowest tail is not
enriched (7 of the lowest 20 are contested, against a 38.9\% base rate).
Across the full 1,796-verse corpus the direction reverses and contested
verses shift left (0.578 versus 0.593), but that observation concerns
material outside the verified scope. The instrument is accordingly used
only as a review queue: low similarity flags verses where translation and
commentary may have drifted apart, for editorial attention first. The
queue's head is informative in practice; its lowest-ranked verse (1.74,
similarity 0.301) is a contested verse whose commentary is dominated by a
doxographic dispute the translation cannot carry.

\begin{figure}[t]
\centering
\begin{tikzpicture}
\begin{axis}[width=0.86\textwidth, height=5.6cm, ybar interval,
  xmin=0.3, xmax=0.8, xlabel={translation--commentary cosine similarity (Book I)},
  ylabel={verses}, legend style={font=\small, at={(0.02,0.95)}, anchor=north west},
  tick label style={font=\small}, label style={font=\small}]
\addplot+[fill=blue!25, draw=datacol] coordinates {
(0.325,3) (0.375,2) (0.425,8) (0.475,21) (0.525,32) (0.575,35) (0.625,29)
(0.675,11) (0.725,3) (0.775,0)};
\addplot+[fill=red!45, draw=gatecol] coordinates {
(0.325,2) (0.375,0) (0.425,2) (0.475,8) (0.525,12) (0.575,11) (0.625,12)
(0.675,6) (0.725,3) (0.775,0)};
\legend{all 144 verses, contested (56)}
\end{axis}
\end{tikzpicture}
\caption{Embedding audit over the Brahmak\={a}\d{n}\d{d}a: distribution of
translation--commentary similarity, with the contested subset overlaid. At
book scale the contested flag does not separate; the low tail serves as a
review queue.}
\label{fig:audit}
\end{figure}

\section{Adversarial Evaluation}
\label{sec:adversarial}

The strongest test of a verification claim is a hostile attempt to break it.
We ran one as a structured red-team: six independent reviewers, each given
only the artifacts and a distinct attacking lens (logical triviality,
fabrication, kernel soundness, completeness, philological faithfulness, and
paper-versus-repository), instructed to falsify rather than confirm; every
serious finding was then reproduced by a separate verifier before being
acted on. We summarize both what held and what did not, because a defense
is only as credible as the attacks it survived.

\paragraph{Attacks that failed.} No \texttt{sorry} or \texttt{admit}, and no
proof-faking construct (\texttt{axiom}, \texttt{native\_decide},
\texttt{implemented\_by}), appears anywhere in the kernel; \texttt{lake
build} succeeds from a clean checkout. Every headline count reproduced from
the artifacts. An independent re-run of the citation gate found all 624
axiom/denial cites and all 450 accepted-reading cites present verbatim in
their sources. Relations were confirmed directional (a reversed relation is
correctly refused), verse namespaces isolated (no cross-verse relation-name
collision), and the Python mirror agreed with the Lean kernel on a battery
of adversarial fragments (empty readings, self-identity, duplicated claims,
a denial equal to an axiom). A swap test---does verse $i$'s accepted reading
pass verse $j$'s contract?---licensed \emph{zero} of all $144 \times 143 =
20{,}592$ ordered pairs, and no rejected reading fails by a dangling entity
identifier (all 361 fail on genuine sort or entailment grounds).

\paragraph{The tautology objection, made falsifiable.} A correct observation
is that each accepted reading is, by construction, a subset of the axioms the
author also wrote (48 of 144 are exact matches), so \texttt{accepted\_adequate}
verifies internal coherence of the author's choices rather than their
independent correctness---exactly the boundary of \S\ref{sec:boundary}. The
objection has force only if the contracts lack discriminative content. The
zero-cross-licensing result answers this mechanically and is now a permanent
regression gate: every accepted reading is licensed by its own contract and
no other. A contract edit that made two verses interchangeable would trip it.

\paragraph{The finding that landed: doxographic conflation.} The one major
finding confirmed on verification was philological, and it is the sharpest
limit of a verbatim-citation gate: the gate cannot distinguish
\emph{endorsing} a claim from \emph{reporting} it. Nine verses in the
doxographic stretch had encoded reported rival views as endorsed axioms. The
fix is the \texttt{reported}-stance mechanism of \S\ref{sec:stance}, which
makes the distinction machine-visible without pretending the machine can
judge it. Two lesser findings---a single contract that both licensed and
denied a claim (1.127), and six contracts with only one refuted reading---were
closed by new validator gates (axiom/denial disjointness; a two-refutation
floor) and by deepening the thin contracts. All fixes are gated and
regression-tested; the review, its verified findings, and the resulting
upgrades are recorded in the ledger.

\section{Discussion}
\label{sec:discussion}

\paragraph{What mechanization buys.} Three things, on the evidence of this
build. \emph{Regression}: interpretive commitments, once contracted, are
protected. An edit to verse 1.46's treatment of the inner word that
contradicts 1.1's identity axioms breaks the build. \emph{Explicitness}: the
contract format forces decisions that prose translation lets one defer (is
\emph{veda} here the corpus or the eternal seed-order?), and the homonym
registry is a by-product catalogue of exactly those decisions.
\emph{Adversarial coverage}: requiring every contract to carry provably
failing rival readings institutionalizes the question of what it would mean
to get each verse wrong. Translators ask that question irregularly;
reviewers cannot ask it at scale.

\paragraph{Relation to the companion programmes.} The loop's discipline
(propose, gate, ledger, adjudicate) is shared with Active Circuit Discovery
and Prabodha \citep{sathish2026acd, sathish2026prabodha}. The instructive
difference is the character of the gate. A statistical gate bounds the
probability of a wrong acceptance; a type-checking gate makes a wrong
acceptance a compile error. The episodes of \S\ref{sec:loop} suggest that
when LLM agents author formal content, the strictest available gate should
sit at the end of the pipeline even when cheaper mirrors exist, because the
mirrors drift (the identifier-hygiene episode) and the agents degrade (the
fabricated-citation episode).

\paragraph{Limitations.} The entailment relation is deliberately shallow
(closure of literal claims under identity symmetry), so adequacy is
membership-like rather than inferential. A richer relation, for example
transitive sambandha chains or defeasible generics, would strengthen the
theorems at the cost of decidability discipline. Contracts formalize the
formalizable core of a verse, not its full rhetorical content. The six-sort
ontology is coarse; persons, for instance, are shoehorned (the child of
1.121 and 1.151 as epistemic subject). The translation and commentary carry
the interpretive standpoint of a single primary source
\citep{iyer1965vakyapadiya}, and the contracts inherit it. Contracts keyed
to rival commentaries (V\d{r}\d{s}abha, Pu\d{n}yar\={a}ja) would turn
inter-commentarial disagreement into diffable, checkable data, and are the
most interesting extension. Finally, the audit instrument is a screen, not
a measure of fidelity, and \S\ref{sec:audit} reports where it fails to
discriminate.

\paragraph{On agents.} The swarm wrote in hours what would take a solo
formalizer months, and its failures were institutionally useful: every class
of agent error that occurred is now a mechanical gate. The suggested general
pattern for LLM-assisted scholarship is to design so that the operative
question is never whether the model hallucinated, but whether the artifact
passed the gate.

\section{Conclusion}
\label{sec:conclusion}

This paper introduced commentary-driven translation validation and
delivered a machine-verified translation layer for the complete
Brahmak\={a}\d{n}\d{d}a of the \emph{V\={a}kyapad\={\i}ya}: 144 adequacy
theorems and 361 compiled refutations over a checksummed edition, produced
by a gated agentic loop whose every acceptance is replayable from an
append-only ledger, and hardened against a structured adversarial review. The method treats the prover not as a reader of
Sanskrit but as a keeper of commitments. Bhart\d{r}hari holds that the
sentence's meaning is an undivided whole, merely manifested by its parts; a
verifier that refuses to let the parts of a translation contradict the
whole of its commentary is in that spirit. Books~II and~III, richer
entailment, and multi-commentary contracts are open work.

\paragraph{Availability.} Code, corpus, contracts, kernel, proofs, ledger,
and the living edition: \url{https://github.com/SharathSPhD/vakya-vallari};
edition mirrors at \url{https://vakya-vallari.vercel.app} and
\url{https://huggingface.co/spaces/qbz506/vakya-vallari}.

\bibliographystyle{plainnat}
\bibliography{refs}

\end{document}
